\section{Natural semantics}\label{sec:bigstep}

In this section we define a natural (i.e.\@ big-step) semantics for
evaluation of expressions and matchings.  The semantics is based in
Sestof's revision~\cite{sestof_1997} of Launchbury's semantics for
lazy evaluation~\cite{launchbury_1993}.

\begin{figure}
  \[
    \begin{array}{rcll}
      e & &     \text{expressions} \\
      e & ::= &  x  & \text{variable} \\
      & | &  \app{e}{y} & \text{function application} \\
      & | & \abstr{m} & \text{matching abstraction} \\
      & | & \constr{c}{y_1,\ldots,y_n} & \text{constructor application} \\
      & | & \llet{\{x_i=e_i\}}{e'} & \text{let(rec) bindings} \\[2ex]
      %%
      m &  & \text{matchings} \\
      m & ::= & \matchreturn{e} & \text{return expression}\\
      & | & \matchfail & \text{failure} \\
      & | & \matchpat{p}{m} & \text{match pattern} \\
      & | & \matcharg{y}{m} & \text{argument supply}\\
      & | & \matchalt{m_1}{m_2} & \text{alternative}\\[2ex]
      %%
      p & & \text{patterns (as before)}
  \end{array}
\]
\caption{Normalized syntax for \lambdaPMC.}\label{fig:syntax-normalized}
\end{figure}

\subsection{Normalized syntax}
As in Launchbury's semantics, we will restrict arguments of
applications to be variables; complex arguments must be explicitly
named using let bindings.  This ensures that we can update a result
after evaluation and properly implemement call-by-need (i.e. lazy)
evaluation.

Figure~\ref{fig:syntax-normalized} defines the syntax for normalized
\lambdaPMC. In the remaining of this paper we assume all expressions
and matchings to be normalized.

\subsection{Preliminary definitions}

Let \emph{heaps} $\Gamma, \Delta, \Theta$ to be finite mappings from
variables to (possibly unevaluated) expressions.  The notation
$\Gamma[\loc \mapsto e]$ in a conclusion extends a heap with an extra
entry; dually, this is also used in hypotheses to extract an entry
from a heap.

Because matchings can encode multi-argument functions and also
argument applications, we have to define the results of evaluations
accordingly. First we define a syntatical measure \arity{m} of the \emph{arity}
of a matching:
\[
  \begin{array}{rcll}
    \arity{\matchreturn{e}} &=&  \arity{\matchfail} &= 0\\
    \arity{(\matchpat{p}{m})} &=& 1 + \arity{m} \\
    \arity{(\matcharg{y}{m})} &=& \max(0, \arity{m} - 1) \\
%%    \arity{{(\matchguard{e}{\constr{c}{\vec{p}}}{m})}} &=& \arity{m} \\
     \arity{(\matchalt{m_1}{m_2})} &=& \arity{m_1} &= \arity{m_2} \\
                            & \multicolumn{2}{l}{\text{(must have equal arity)}}
  \end{array}
\]
The last condition generalizes the Haskell syntax rule that the number
of patterns for each clause of a definition must be the
same~(Section~4.4.3.1 of~\cite{haskell_2010_report}).

An expression is in weak head normal normal (whnf) $\whnf$ if it is either
a matching abstraction of arity greater than zero or a constructor:
\[
\begin{array}{rcll}
  \whnf  &::=&  \lambda m & \text{such that}~ \arity{m}> 0 \\
     &|& \constr{c}{y_1,\ldots,y_n}
\end{array}
\]
If $\arity{m}>0$ then $m$ expects at least one argument, i.e.\@
behaves like a lambda abstraction.  If $\arity m= 0$ then $m$ is
\emph{satured} (i.e.\@ fully applied) and therefore \emph{not} a whnf.

This definition of weak normal forms implies that partially applied
matchings will not be evaluated e.g.\@
$\lambda(\matcharg{z}{\matchpat{x}{\matchpat{y}{\matchreturn{e}}}})$
is in whnf. This is similar to a partial application in an abstract
machine with multi-argument functions e.g.\@ the
STG~\cite{jones_1992}.  However, in our case this is done not for
efficency but rather to simplify evaluation by ensuring that we only
ever evaluate saturated matchings.

We also define a weak normal form $\matchresult$ for matching results, namely,
either a return expression or a failure:
\[
  \matchresult \quad::=\quad \matchreturn{e} \quad|\quad \matchfail
\]
  


\begin{figure}
  \begin{gather*}
    \framebox{\ensuremath{\Gamma;\lset;e \expev \Delta; w}} \qquad \text{expression evaluation} \\[4ex]    
    %
    %
    \prooftree
    \justifies
    \Gamma; \lset; w \expev \Gamma; w
    \using{\bigrule{Whnf}}
    \endprooftree\\[3ex]
    %
    \prooftree
    \arity{m}= 0 \and
    \Gamma; \lset; [\,];\, m \matev \Delta;\matchreturn{e} \and
    \Delta; \lset; e \expev \Theta;\,w
    \justifies
    \Gamma;\lset; \lambda m \expev \Theta;w
    \using{\bigrule{Sat}}
    \endprooftree\\[3ex]
    %    %
    \prooftree
    \Gamma;\lset\cup\{\loc\}; e \expev \Delta; w 
    \justifies
    \Gamma[\loc\mapsto e]; \lset; \loc \expev \Delta[\loc\mapsto w]; w
    \using{\bigrule{Var}}
    \endprooftree \\[3ex]
    %% 
    \prooftree
    \Gamma;\lset; e\expev \Delta;\lambda m
    \and
    \Delta;\lset; \lambda (\matcharg{\loc}{m}) \expev \Theta;w 
    \justifies
    \Gamma;\lset; (e~\loc) \expev \Theta; w
    \using{\bigrule{App}}
    \endprooftree \\[3ex]
    %
    %
    \prooftree
    \Gamma[\{\loc_i\mapsto \widehat{e}_i\}];\lset; \widehat{e}' \expev \Delta;w
    \justifies
    \Gamma;\lset; \llet{\{x_i=e_i\}}{e'} \expev \Delta;w
    \using{\bigrule{Let}} 
    \endprooftree \\
    \text{where}~\loc_i ~\text{are fresh}  \\%% ~\lset, \Gamma, \llet{\{x_i=e_i\}}{e'} \\
    \widehat{e}_i = e_i[\loc_1/x_1,\ldots,\loc_n/x_n], \\
    \widehat{e}' = e' [\loc_1/x_1,\ldots,\loc_n/x_n]
  \end{gather*}
  \caption{Expression evaluation}\label{fig:expr-eval}
\end{figure}

\begin{figure}
  \begin{gather*}
    %
    \framebox{\ensuremath{\Gamma;\lset;A;m\matev \Delta;\matchresult}} \qquad
    \text{matching evaluation} \\[4ex]
    %
    \prooftree
    \justifies
    \Gamma; \lset; \argstack; \matchreturn{e} \matev \Gamma; \matchreturn{\matcharg{\argstack}{e}}
    \using{\bigrule{Return}}
    \endprooftree \\[3ex] 
    %
    \prooftree
    \justifies
    \Gamma; \lset; A; \matchfail \matev \Gamma;\matchfail
    \using{\bigrule{Fail}}
    \endprooftree\\[3ex]
    %
    \prooftree
    \Gamma; \lset; (y:\argstack); m \matev \Delta;\matchresult
    \justifies
    \Gamma; \lset; \argstack; \matcharg{y}{m} \matev \Delta; \matchresult
    \using{\bigrule{Arg}}
    \endprooftree\\[3ex]
    %
    \prooftree
    \Gamma;\lset; \argstack; m[y/x] \matev \Delta; \matchresult
    \justifies
    \Gamma; \lset; (y:\argstack); \matchpat{x}{m} \matev \Delta; \matchresult
    \using{\bigrule{Bind}}
    \endprooftree \\[3ex]
    %
    \prooftree
    \begin{array}{l}
    \Gamma;\lset; y\expev \Delta; \constr{c}{y_1,\ldots,y_n} \\
    \Delta;\lset;\argstack; \matcharg{y_1}{p_1} \amatchpat \ldots \amatchpat \matcharg{y_n}{p_n} \amatchpat m
      \matev \Theta;\matchresult
    \end{array}
    \justifies
    \Gamma; \lset; (y:\argstack); \matchpat{\constr{c}{p_1,\ldots,p_n}}{m}
    \matev \Theta;\matchresult
    \using{\bigrule{Cons1}}
    \endprooftree \\[3ex]
    %
    %
    \prooftree
    \Gamma;\lset; y\expev \Delta; \constr{c'}{y_1,\ldots, y_k} \qquad
    c \neq c' \lor n \neq k
    \justifies
    \Gamma; \lset; (y:\argstack); \matchpat{\constr{c}{p_1,\ldots, p_n}}{m}
    \matev \Delta; \matchfail
    \using{\bigrule{Cons2}}
    \endprooftree    \\[3ex]
    %
    \prooftree
    \Gamma;\lset;\argstack; m_1 \matev \Delta;\matchreturn{e}
    \justifies
    \Gamma;\lset;\argstack; (\matchalt{m_1}{m_2}) \matev \Delta;\matchreturn{e}
    \using{\bigrule{Alt1}}
    \endprooftree \\[3ex]
    %
    %
    \prooftree
    \Gamma;\lset;\argstack; m_1 \matev \Delta;\matchfail  \qquad
    \Delta;\lset;\argstack; m_2 \matev \Theta; \matchresult
    \justifies
    \Gamma;\lset;\argstack; (\matchalt{m_1}{m_2}) \matev \Theta;\matchresult
    \using{\bigrule{Alt2}}
    \endprooftree
  \end{gather*}
  
  \caption{Basic matching evaluation}\label{fig:match-eval}
\end{figure}




\subsection{Evaluation rules}

Evaluation is defined in Figures~\ref{fig:expr-eval}
and~\ref{fig:match-eval} by two mutually recursive judgments:
\begin{description}
  \item[$\Gamma;\lset;e \expev \Delta;w$]  Evaluating 
   expression $e$ from heap $\Gamma$ yields heap $\Delta$ and result $w$;
  \item[$\Gamma;\lset;\argstack; m \matev \Delta;\matchresult$] 
    Evaluating matching $m$ from heap $\Gamma$ and argument stack \argstack\
    yields heap $\Delta$ and result \matchresult.
  \end{description}

  In the $\matev$ judgements the argument stack \argstack\ is
  a sequence of variables representing the pending arguments
  to be applied to the matching expression.

  In both judgments the set $\lset$ keeps track of variables under
  evaluation and is used to ensure freshness of variables in the
  \bigrule{Let} rule~\cite{sestof_1997}.
  \begin{definition}[Freshness condition]
    A variable $y$ is fresh if it does not occur (free or bound) in
    $L$, $\Gamma$ or $\llet{\{x_i=e_i\}}{e}$.
  \end{definition}
  

\paragraph{Remarks about rules for  expressions (Figure~\ref{fig:expr-eval}).}
  
  Rule~\bigrule{Whnf} terminates evaluation immediately when we reach a whnf,
  namely a non-saturated matching abstraction or a constructor.
  
  Rule~\bigrule{Sat} applies to saturated matching abstractions;
  if the matching evaluation succeeds then
  we proceed to evaluate the expression returned.
    
  Rule~\bigrule{Var} forces evaluation of an expression in the heap; as in
  Launchbury and Sestof's semantics, we remove the entry
  from the heap  while performing the evaluation (``black-holing'')
  and update the heap with the result afterwards.
    
  Rule~\bigrule{App} first evaluates the function to
  obtain a matching abstraction and then evaluates
  the the argument application.
    
  Rule~\bigrule{Let} is identical to the one by Sestof: it allocates
  new expressions in the heap (taking care of renaming) and
  continues evaluating the body of the let expression.
  
  \paragraph{Remarks about rules for matchings (Figure~\ref{fig:match-eval}).}

  Rule~\bigrule{Return} terminates evaluation sucessfully
  when we reach a return expression.
  The notation $A\amatcharg e$ represents the nested applications
  of  left over arguments on the stack $A$ to the expression $e$. The definition is:
  \begin{align*}
  \matcharg{[\,]}{e} &= e \\
  \matcharg{(y:ys)}{e}  &= \matcharg{ys}{(e~y)}
  \end{align*}

  Rule~\bigrule{Fail} terminates evaluation unsuccessfully when
  we reach the match failure \matchfail.

  Rule~\bigrule{Arg} pushes an argument onto the stack and carries
  on evaluation.
  
  Rule~\bigrule{Bind} binds a variable pattern to
  an argument on the stack. This simply a renaming of the pattern
  variable $x$ to the heap variable $y$.
  
  Rules~\bigrule{Cons1} and \bigrule{Cons2} handle the successful
  and unsucessful matching of a constructor pattern; in the later
  case the matching evaluation returns \matchfail.  Note also that
  \bigrule{Cons1} continues the matching of sub-patterns in left to
  right order.
  

  Rules~\bigrule{Alt1} and \bigrule{Alt2} handle progress and failure
  in alternative matchings.  Note that in \bigrule{Alt2} evaluation
  continues with the updated heap $\Delta$ because the evaluations
  effects of the failed match need to be shared when evaluating $m_2$.
  Note also that the argument stack is shared between $m_1$ and $m_2$
  implementing rule $(\amatcharg\mid)$ of Section~\ref{sec:reduction}.
  This sharing justifies why we restrict argument supply to single
  variables: allowing arbitrary expressions as arguments could
  duplicate computation because the same expression could be evaluated
  in more than one alternative. By restricting to single variables
  we can ensure that results are shared.
    
    
\subsection{Dealing with pattern guards and view patterns}\label{sec:pattern-guards}

The syntax of matchings in~Figure~\ref{fig:syntax-normalized} and
evaluation rules of Figure~\ref{fig:match-eval} deal only with
argument matches of the form $\matcharg{y}{m}$.  This is done to
allow sharing required by lazy evaluation but begs the question
of how to encode the boolean and pattern guards. 

Consider again the \textit{nodups} example of
Section~\ref{sec:examples}; at first sight, it may seem that we can
use let to introduce a name for the boolean guard $(x==y)$:
\[
  nodups = \lambda \!\!\begin{array}[t]{l}
    ((x:xs) \amatchpat xs \amatcharg (y:xs') \amatchpat \\
     \qquad \matchreturn{\alet~ b = (x==y)~ \ain~
             \lambda(b \amatcharg \textsf{True} \amatchpat \matchreturn{nodups~ xs})}  \\
                         \mid \ldots~)
                       \end{array}
\]
However, this does not work: if $b$ evaluates to \cons{False},
then $\lambda(b \amatcharg \textsf{True} \amatchpat \ldots)$ gets ``stuck''
as $\lambda \matchfail$ and the subsequent equations are never attempted.

One solution would be to join the two alternatives for the boolean guard:
\[
  nodups = \lambda \!\!\begin{array}[t]{l}
    ((x:xs) \amatchpat xs \amatcharg (y:xs') \amatchpat \\
     \quad \lreturn\alet~ b = (x==y)~ \ain~
                         \lambda(b \amatcharg \!\!\!\begin{array}[t]{l}
                                                (\cons{True} \amatchpat \matchreturn{nodups~ xs} \\
                                                \mid \cons{False} \amatchpat \matchreturn{x:nodups ~ xs})\rreturn\end{array}  \\
                         \mid [\,] \amatchpat \matchreturn{[\,]}
                       \end{array}
\]
The disadvantage of this encoding (besides verbosity) is that we lose
the direct connection with the original three equations of the source
program.

A better solution is to allow pattern guards explicitly in the
normalized syntax for \lambdaPMC:
\[
  \begin{array}{llll}
  m &::=& \ldots & \text{(as in Figure~\ref{fig:syntax-normalized})} \\
    & | & \matcharg{e}{\matchpat{\constr{c}{\vec{p}}}{m}} & \text{pattern guards}
                                                            \end {array}
\]

Now \textit{nodups} can be written exactly as in Section~\ref{sec:examples}:
\[
  nodups = \lambda
  \!\!\begin{array}[t]{l}
    ((x:xs) \amatchpat
    xs \amatcharg (y:xs') \amatchpat \\
     \qquad\qquad (x==y) \amatcharg \textsf{True} \amatchpat \matchreturn{nodups~ xs}  \\
    \amatchalt (x:xs) \amatchpat  \matchreturn{x: nodups~xs} \\
    \amatchalt [\,] \amatchpat \matchreturn{[\,]})
  \end{array}
\]

The extra evaluation rules need to deal with pattern guards
are analogous to \bigrule{Cons1} and \bigrule{Cons2}:
\begin{gather*}
  \prooftree
    \begin{array}{l}
    \Gamma;\lset; e\expev \Delta; \constr{c}{y_1,\ldots,y_n} \\
    \Delta;\lset;\argstack; \matcharg{y_1}{p_1} \amatchpat \ldots \amatchpat \matcharg{y_n}{p_n} \amatchpat m
      \matev \Theta;\matchresult
    \end{array}
    \justifies
    \Gamma; \lset; \argstack; \matcharg{e}{\matchpat{\constr{c}{p_1,\ldots,p_n}}{m}}
    \matev \Theta;\matchresult
    \using{\bigrule{Guard1}}
    \endprooftree \\[3ex]
    %
    %
    \prooftree
    \Gamma;\lset; e\expev \Delta; \constr{c'}{y_1,\ldots, y_k} \qquad
    c \neq c' \lor n \neq k
    \justifies
    \Gamma; \lset; \argstack; \matcharg{e}{\matchpat{\constr{c}{p_1,\ldots, p_n}}{m}}
    \matev \Delta; \matchfail
    \using{\bigrule{Guard2}}
    \endprooftree
  \end{gather*}
%
Note that when $e$ is a single variable then these rules
produce the same effect as the combination of \bigrule{Arg} 
with \bigrule{Cons1} or \bigrule{Cons2}.

As seen in Section~\ref{sec:examples}, we can use pattern guards for
encoding view patterns. Recall 
the \textit{length} function over join lists:
\[
  length = \lambda \!\!\begin{array}[t]{l}
                     (t \amatchpat view~t \amatcharg 
      \cons{Nil} \amatchpat \matchreturn {0} \\
      \mid t \amatchpat view~t \amatcharg \cons{Cons}(h,xs) \amatchpat \matchreturn{1+length~ xs})
    \end{array} 
\]
One issue with this encoding is that evaluation of the view function is not shared.
By constrast, GHC will try to optimize calls to view functions in
such cases~\cite{ghc_guide_view_patterns}.
Such as optimization can be expressed in \lambdaPMC\ by using a let to explicily share the result.
However, we lose the one-to-one relation with the source program:
\[
  length' = \lambda (t \amatchpat\!\! \begin{array}[t]{l}
                                    \lreturn \alet~ v = view~t \\
  ~\ain~ \lambda (v \amatcharg
  \!\!\!\begin{array}[t]{l}
    (\cons{Nil} \amatchpat \matchreturn {0}  \\
    \mid \cons{Cons}(h,xs) \amatchpat \matchreturn{1+length'~ xs}) \rreturn
  \end{array}
                                  \end{array}
\]

    

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
